
\section{On Withdrawal from a Synod}

Editorial article in "Folkebladet" for February 23, 1898.



We included Pastor A. Wright’s piece on "Going out" in the last issue of "Folkebladet."\footnote{
Pastor Wright’s piece reads as follows:

Should we not "go out" again?
A Letter to My Friends.

[Reprinted in Folkebladet for February 16, 1898.]

Among those who first went from Augsburg Seminary to Gjertsen’s Church at the Union meeting in 1890 was the undersigned. The reasons for this hesitation were several, which shall not be mentioned here; at any rate not this time. Since that time, the development in our Synod has [also] shown that something of what I foresaw would happen has in fact taken place. Among those of you who have looked critically at one thing and another, I am one; and because of that, some radicals have lost confidence in me, insofar as they ever had any, perhaps chiefly because I have defended the Augsburg people in one thing and another (never in one and all things), when they themselves were not present, and some of them count themselves among my personal friends. I have sought not to do them harm, but [to show] good.

So then, when I have found in our Synod things with which I was not in agreement, and have spoken to some of the leaders about this, without their having, so far as I can see, followed all my suggestions, would it not then have been more right to leave the Synod and seek to bring my friends with me, so that we could be free and work freely? It is about this that I now, as I "request the floor" in "Lutheraneren," should like to be allowed to include something in this letter.

1. God wills that His Church should be one, not only in the Spirit, but also in the body. John 17. 1 Cor. "Frailties" among persons in the Synod do not annul this will of God. That this will should be done, toward that end and upon that the Christian works as well as he can, when he understands his task and is obedient.

2. Sins and faults among fellow Christians are to be punished and disciplined. Matt. 18. There is much good material in the Synod for the building of the Synod. If I am worth anything as a workman, I understand how to value and work it in company with "fellow workers," so that we may build upon God’s building; but this we cannot do by rejecting and being impatient.

3. How many years we are to "forbear" and labor before we give up hope of improvement is not prescribed in God’s Word; in the vast majority of cases it is the whole lifetime.

4. I have the right to go out of the congregation or the Synod only:
   a) When false doctrine is taught and defended.
   b) When, on that account, false practice (ungodly conduct) develops.
   c) When testimony against these things (the testimony of Scripture) is stamped as error and is thus forbi[?]dden from being preached.

5. To go out is the very highest degree of church discipline. Before I can resort to this means, I must in my conscience before God be certain that the one or those from whom I separate myself are as "a Gentile or a tax collector."

6. If I myself sin (commit wrong) while I am to discipline, I become a partaker in the guilt and therefore have one support less for my case.
   "Rarely does only one of two cause a quarrel."

7. By taking note of the withdrawals that have not taken place on the basis described in God’s Word (mentioned above), I have, in the years I have been here in this country, seen that such withdrawal ends badly. There is danger when you button your vest and make a mistake the first time; it goes well for quite a long while. But in the end you make the discovery that you have either one buttonhole too many or one button too many. And in any case, something essential is either too much or too little after a fair lapse of time, where withdrawals have been undertaken for invalid reasons.

Therefore, dear friends, let us not go out! Shall we flee the house when there is danger? What sort of watchmen are we then? No. Let us labor honestly and openly against that which, according to our conviction, is wrong, but not run away. We also need to be tested. There are, after all, two possibilities: what we think is wrong may be wrong, but we ourselves might also be mistaken. Perhaps we do not see it until we have come a good way farther on.

As for me, I may say: late I went in; late I go out. So before that, entirely different grounds than those now on the agenda will be required before I can entertain such thoughts.

Brotherly greeting from yours, bound together in the Lord,

A. Wright.}

And we should very much like to ask many to read it again aloud. It contains a good deal that is sound, if it is applied to withdrawal from a congregation; at the same time, it is extraordinarily weak when it is applied to withdrawal from a "Synod."

We also wish to make a few remarks on the piece, which we hope will bring Christian people onto the right track in this matter.

Concerning the reasons why one ought not to go out of "our Synod," Pastor Wright says among other things:

"1. God wills that His Church should be one, not only in the Spirit but also in the body. John 17. 1 Cor. Frailties among persons in the Synod do not annul this will of God."

It appears that Pastor Wright means that Jesus in John 17, the so-called High-Priestly Prayer, prayed about and for an outward organization of the Church, so that this prayer is fulfilled by all congregations joining themselves to one and the same outward organization, and then no doubt preferably the United Church.

Here, then, the question arises:
Is the Church’s inward and outward unity advanced or hindered by the formation of the so-called "Synods"?

We know from the New Testament that the Church’s unity is not hindered by the many congregations. Jesus and His apostles were not opposed to the number of congregations increasing in a lawful manner.

But if synod after synod is formed, and synod against synod, and each of them claims to be the one true Church, does not this hinder the Church’s unity?

Is it absolutely certain that the formation of these governing church bodies really belonged to the plan of Jesus and the apostles?

But if it should be so that the freedom and independence of the congregations, without any governing "Synod" over them, is the original order, the one best fitted for the Church’s unity, then it is indeed asked whether those who tighten the synodical bond the most and make it into such a matter of conscience also serve the Church’s unity best.

For we all know that congregations are divided and torn apart on account of the so-called synods; is this then also an advancement of the Church’s unity?

And Pastor Wright further says:
"4. To go out of the congregation or the Synod I have a right only" etc.

To this we would only remark that this view, according to which congregation or synod is placed absolutely side by side, seems to us either to place the congregation too low or the synod too high.

"Folkebladet" does not wish to pronounce itself on the many reasons that speak for withdrawal from the United Church. We are certain that the congregations will find out where the shoe pinches. If there are any among them who feel that they cannot with good conscience remain with the "Synod," then they know well enough what they are to do. "Folkebladet" has no anxiety about the merely outward side of the matter.

Georg Sverdrup




