\section*{Project Disclaimer}

This volume is an independent scholarly and technical project. It is not affiliated with, endorsed by, or issued under the authority of any church body, academic institution, or other organization, including but not limited to the American Lutheran Free Church (AFLC), the Sverdrup Society, Augsburg Seminary, or any related bodies.


The texts presented herein are historical documents. The English translations, transcriptions, and editorial materials are supplied solely for educational and research purposes. They do not claim doctrinal, ecclesiastical, institutional, or critical authority.

The English translations in this volume were produced with the aid of artificial intelligence–based language tools operating under detailed scholarly instructions. Editorial responsibility in the present edition is limited to methodological design, critical oversight, source comparison, and presentation.

The original source texts remain definitive.

Further information regarding scope, methodology, limitations, and revision history may be found in \texttt{DISCLAIMER.txt} and in the project repository on GitHub:

\url{https://github.com/apdahlen/Guidance}


\section*{Editorial Note}

This document is a working translation and reconstruction from a source that is in poor physical and typographic condition. Some passages remain uncertain owing to faded print, damaged pages, archaic spelling, irregular orthography, and the inherent limitations of OCR and other machine-assisted recovery methods; such tools can assist transcription, but they cannot by themselves guarantee correct readings in difficult witnesses.

Readers should therefore treat this as a provisional scholarly working text rather than a final critical edition. The project is maintained openly on GitHub and constructed in \LaTeX, so that the text, formatting, annotations, and editorial decisions may be revised transparently over time.

Corrections, source checks, and suggested improvements are welcome through the project repository on GitHub, where revisions and editorial decisions may be tracked transparently.


\section*{Future Work}

\begin{itemize}
  \item AI-assisted batch analysis of biblical quotations to favor DAN 1819 / 1871 over KJV
  \item Further editorial work to improve consistency across the volume.
  \item Continued refinement to increase fidelity to the Danish without sacrificing readable English.
\end{itemize}
