
\section{The Free Church Came}

Editorial article in "Folkebladet," June 23, 1897.

It was not any particularly great event. It was nothing other than to establish what is already being practiced, and to give it a name. We do not expect anything great from the name, and we know that if the Free Church is ever to become anything, long and faithful work will be needed all the while.

The question therefore is really this: Will the work come that is designated by the name the Free Church?

People always say that the people are not mature for freedom; therefore they must not have it. In this way they say: The congregations are not mature for freedom; therefore they must not have it.

The truth is that one ought to reason the other way: If the congregations are not mature for freedom, then let them take freedom at once, so that they may become mature.

The congregations cannot properly receive freedom; they can only take it. That is to say, they can receive it only from God; for surely there is no human being who can give it to them. Therefore we say that the sooner the congregations cease asking human beings for permission, and take the freedom that lies ready for them in the Word, the better.

That is the Free Church. When the congregations begin, each of them, to order themselves after the apostolic pattern, so that in all their work they seek only and follow only the light of the Word and the Spirit, then with that the free-church work has also begun.

But the congregations, or the single congregation, cannot begin anything, unless some man or woman in the congregation begins.

Therefore there are many who have it in such a way that they turn this matter upside down. They go about waiting for a congregational resolution before they can begin to work. If there is something that requires a congregational resolution, then some man or woman must begin to work in order to get the congregational resolution adopted; that is clear enough. The resolution comes after a proposal; if no one makes the proposal, then the resolution does not come into being either.

But by far the greater number of Christian actions can and must be done by men and women without a congregational resolution. And there the free-church work begins in the strictest sense. It begins in the believers’ work for the salvation of souls.

Therefore this, that "the Free Church came," is so insignificant. But this is the great question, whether the Free Church will come. That depends upon each individual Christian man and woman throughout the congregations.

But surely this is not anything new? All who are against the Free Church will answer No. The Free Church people feel that to this they must answer both No and Yes. We mean that free Christian work is as old as the time of the apostles. But we also mean that it has grown very quiet around the believers’ free Christian work since the state church ordered and commanded, forbade and prescribed, everything by law. We have seen that in the so-called "societies" free Christian work has almost ceased; in some places, indeed, it has in many ways even been forbidden, because such a work of freedom would disturb the generally prevailing sleep. Priest and sexton and collection committees are to work; others ought simply to leave it alone.

If the believers will not stir themselves more and more as time goes on, then the Free Church is in vain. Therefore the name is only a hope that has been uttered, a call that goes out to the Christians in all congregations.

The Free Church came; will it truly come?

Georg Sverdrup
