
\section{Freedom and Equality}

Editorial article in “Folkebladet” for October 27, 1897.

“Lutheraneren” has had to abandon the hope of proving that the congregations have lost their freedom in the Free Church. But now it has discovered another danger: equality is gone. Congregations which came out from under the church body’s guardianship are no longer equal.

Why not? Because the Free Church does not have the representative system.

So then, because the congregations have taken back the power and authority of which the representative system deprives them, they therefore cannot be equal.

It is said: Yes, but at the Free Church’s meetings there may be so many from one congregation that the other congregations are wronged.

This would have some value if our meetings were representative, with supremacy over the congregations. Since they have no other influence than the purely voluntary, and since they may be held anywhere, or wherever it may be, there is surely no sense in saying that they make the congregations unequal.

The Free Church consists of congregations, not of annual meetings. Each congregation by itself governs itself and all its affairs without asking anyone else. If that is not equality, then it is surely difficult to find it.

But the equality which consists in this, that all congregations resemble one another as one drop of water another, that they are equally large, equally spiritual, equally wise, equally gifted, equally godly, equally useful, etc., etc., endlessly—that kind of equality has surely never existed and can hardly exist either.

But this equality, that they are equally free, equally responsible, equally precious, equal “before the Law,” that kind of equality there is surely more of when one is free of the representative system than when one lies under it....

Georg Sverdrup
