% 04 Apr: AI clean

\section{On Inner Mission}

"Folkebladet," November 22, 1893.

Our starting point is this: that Inner Mission is simply the proclamation of the Word and the administration of the Sacraments for the founding and upbuilding of congregations in places where the people are either too poor or too indifferent to support the church’s work themselves. We do not regard mission work primarily as an expansion of the Association, but as the forming of new Lutheran congregations. We do not regard the mission contributions as ordinary expenses of the Association, which is why they are not paid out of the Association’s treasury either. We do not regard the mission gifts as an instrument of power in the hands of a mere majority, so that it may use these holy gifts according to the majority’s whim. We do not regard the mission committee either as a part of the governing board of the Association or as an independent mission board, but as a servant intermediary between the congregations that give and the congregations that receive help. We therefore believe that faithfulness to the Association in mission matters is the greatest possible zeal for the salvation of souls, for the upbuilding of Lutheran congregations, for the diligent proclamation of the Word, and for the right administration of the Sacraments.

It might therefore very well be an act of great faithfulness in mission work even if we were to help a congregation belonging to Hauge’s Synod or to the Norwegian Synod in a place where it was truly needed. The more genuine Lutheran impartiality can be shown in Inner Mission, the better for our people, for our church, and for the Association. . . .

The minority’s mission committee asks no other question of those it supports than this: Is there need, and has help been denied by the majority? Whether the congregations we support are willing to promise aid to Augsburg or not, we have not asked them. We know that the proclamation of the Word among them would suffer a corresponding curtailment if we did not help them; therefore we help as far as we can.

Georg Sverdrup.

